Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-27 manuscript/manuscript.tex
Minor changes.

\documentclass[11pt, letterpaper]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
% Allow inline lists
\usepackage[inline]{enumitem}
\usepackage[backend=biber, style=authoryear, autocite=inline, isbn=false, doi=false, url=false, eprint=false]{biblatex}
% Fix biblatex behavior that writes "In: " before journal name.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}

@@

\bibliography{references}

@@

\abstract{[NOTE: OLD]
The genetic diversity of a population reflects its demographic and
evolutionary history. Methods for inferring this history typically
assume that the ancestry of a sample can be modeled by the Kingman
coalescent process. A defining feature of the Kingman coalescent is
that it generates genealogies that are binary trees: no more than two
ancestral lineages may coalesce at the same time. However, this
assumption breaks down under several scenarios. For example, pervasive
natural selection, rapid spatial range expansion, and extreme
variation in offspring number can all generate genealogies with
``multiple-merger'' events in which more than two lineages coalesce
instantaneously. Therefore, detecting multiple mergers is important
both for understanding which forces have shaped the diversity of a
population and for avoiding fitting misspecified models to data.
Current methods to detect multiple mergers rely on the average site
frequency spectrum (SFS). However, the signatures of multiple
mergers in the average SFS are also consistent with a Kingman
coalescent process with a time-varying population size. Here, I
present a new method for detecting multiple mergers based on
the mutual information of the joint allele frequency spectrum at pairs of linked sites. Unlike the average SFS,
the mutual information depends mostly on the topologies of genealogies
rather than their branch lengths and is therefore robust to most
demographic effects.}

@@

Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
In particular, we demonstrate that a transformation of the 2-SFS that we refer to as \emph{frequency pointwise mutual information} (fpmi) is sensitive to multiple mergers but relatively invariant under time-varying coalescent rates.
We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
We demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

@@

We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
These files contain the sequence alignments of all flies in the sample on all chromosome arms.
We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order by sample name.
(As a result, the data for each chromosome arm is from a slightly different subset of the individuals.)

@@

We compared the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
We are interested in whether $\fpmi$ can distinguish among coalescent models that produce similar distortions in the SFS.
To this end, we computed a version of Tajima's D \autocite{Tajima1983} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
All results are computed numerically using the results of \cite{Fu1995}, \cite{ZivkovicWiehe2008}, and \cite{BirknerEtAl2013} (see Methods).

@@

After filtering for coverage, removing chromosome arms with known inversions, downsampling to $n=100$ samples per autosomal chromosome arm, and identifying 4-fold degenerate sites, we selected the central region of each chromosome that have consistent high diversity (Methods).
Because the average pairwise diversity varies between arms---possibly reflecting selection or different sets of segregating inversions---we performed all subsequent calculations on each arm independently.
We fit a demographic model to the site frequency spectra of these central regions using \texttt{fastNeutrino} \cite{BhaskarEtal2015}.
We fit a 3-epoch piecewise-constant model, with four free parameters: two changepoints and two population size ratios.
We report our fitted parameters in Table~\ref{tab:DPGP_params}.
We then simulated under our fitted model using \texttt{msprime} and computed the expected and observed SFS, $\fpmi$, and $\hilopmi$ (\fig{fig:dpgp3}, See methods for details.).

>>>>

\documentclass[11pt, letterpaper]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
% Allow inline lists
\usepackage[inline]{enumitem}
\usepackage[
    backend=biber,
    citestyle=authoryear,
    style=authoryear,
    autocite=inline,
    isbn=false, doi=false, url=false, eprint=false,
    uniquename=false,
    maxcitenames=2
]{biblatex}
% Fix biblatex behavior that writes "In: " before journal name.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}

@@

\addbibresource{references.bib}

@@

\abstract{[NOTE: OLD]
The genetic diversity of a population reflects its demographic and evolutionary history. Methods for inferring this history typically assume that the ancestry of a sample can be modeled by the Kingman coalescent process. A defining feature of the Kingman coalescent is that it generates genealogies that are binary trees: no more than two ancestral lineages may coalesce at the same time. However, this assumption breaks down under several scenarios. For example, pervasive natural selection and extreme variation in offspring number can both generate genealogies with ``multiple-merger'' events in which more than two lineages coalesce instantaneously.  Therefore, detecting multiple mergers is important both for understanding which forces have shaped the diversity of a population and for avoiding fitting misspecified models to data. Current methods to detect multiple mergers rely on the site frequency spectrum (SFS). However, the signatures of multiple mergers in the SFS are also consistent with a Kingman coalescent process with a time-varying population size. Here, we present a new method for detecting multiple mergers based on the pointwise mutual information of the two-site frequency spectrum for pairs of linked sites. Unlike the SFS, the pointwise mutual information depends mostly on the topologies of genealogies rather than their branch lengths and is therefore insensitive to population size change.}

@@

Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtAl2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
In particular, we demonstrate that a transformation of the 2-SFS that we refer to as \emph{frequency pointwise mutual information} (fpmi) is sensitive to multiple mergers but relatively invariant under time-varying coalescent rates.
We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
We demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

@@

We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtAl2011}.
The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
These files contain the sequence alignments of all flies in the sample on all chromosome arms.
We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order by sample name.
(As a result, the data for each chromosome arm is from a slightly different subset of the individuals.)

@@

We compared the $\fpmi$ of the Kingman coalescent with and without population growth to the $\fpmi$ of the beta coalescent, for pairs of sites without recombination.
We are interested in whether $\fpmi$ can distinguish among coalescent models that produce similar distortions in the SFS.
To this end, we computed a version of Tajima's D \autocite{Tajima1989} normalized to be invariant under changes in the mean coalescent rate: $D = (\Pi - \hat{\theta}_W) / \Pi$, where $\hat{\theta}_W$ is Watterson's theta, a linear function of the SFS \autocite{Watterson19??}.
Negative values of $D$ indicate an overabundance of low-frequency mutations relative to the time-homogeneous Kingman expectation.
All results are computed numerically using the results of \cite{Fu1995}, \cite{ZivkovicWiehe2008}, and \cite{BirknerEtAl2013} (see Methods).

@@

After filtering for coverage, removing chromosome arms with known inversions, downsampling to $n=100$ samples per autosomal chromosome arm, and identifying 4-fold degenerate sites, we selected the central region of each chromosome that have consistent high diversity (Methods).
Because the average pairwise diversity varies between arms---possibly reflecting selection or different sets of segregating inversions---we performed all subsequent calculations on each arm independently.
We fit a demographic model to the site frequency spectra of these central regions using \texttt{fastNeutrino} \cite{BhaskarEtAl2015}.
We fit a 3-epoch piecewise-constant model, with four free parameters: two changepoints and two population size ratios.
We report our fitted parameters in Table~\ref{tab:DPGP_params}.
We then simulated under our fitted model using \texttt{msprime} and computed the expected and observed SFS, $\fpmi$, and $\hilopmi$ (\fig{fig:dpgp3}, See methods for details.).
